% Abstract for the ESIPS honours thesis.
% Intended to be pulled into main.tex via \input{tex_files/0. Abstract/abstract.tex}.
% Assumes an IEEEtran document; wrapped in the abstract environment so it renders
% in the single-column header before \maketitle's body. Contains no preamble.
%
% Feedback pass (July 2026): rewritten against the supervisor's comments.
% Both paradigms are now defined independently; models and metrics are named
% (GPT-5.4-nano, text-embedding-3-small, Qwen3-Embedding-4B, Cohere Rerank 4.0
% Pro); "joint accuracy" is stated in plain words rather than named; each
% number sits beside the finding it supports; "confined to the two artefacts"
% and "component benchmarks select every deployed model" reworded plainly.
% Feedback pass 2 (July 2026, clarity sweep): a tie-back sentence now closes
% the component results, connecting them to the aim (structured and semantic
% retrieval combine rather than trade off). The final sentence stays a
% placeholder until the system-level evaluation (RQ3) runs. No length limit
% applies per the submission portal.
% Comparator pass (July 2026): the RAG-LLM comparator is removed from the
% design; the final sentence now names the three deployed engines (Google
% Scholar, PubMed, Semantic Scholar). Rationale and protocol: research_design
% III-D6 and the NOTE there.
\begin{abstract}
Deployed literature-search engines force a choice between two retrieval
paradigms. Structured retrieval matches exact terms and typed metadata, so a
paper either satisfies the query or it does not, and the same query always
returns the same answer. Semantic retrieval ranks papers by meaning, so it can
surface relevant papers whose wording differs from the query. This thesis
designs, builds, and evaluates a scholarly search system that combines the
two, built to production standard for Latent Knowledge, an industry partner
whose users are biomedical researchers. A typed knowledge graph of
approximately 150 million biomedical works, ingested from the 2026-02-03
OpenAlex snapshot, anchors a three-stage cascade in which structured graph
retrieval returns the works that provably satisfy the query, dense bi-encoder
retrieval ranks that set by semantic similarity, and a cross-encoder re-ranks
the final results. A language model performs only the two tasks the structured
stage needs, building the per-work keyword index offline and translating each
query into an intermediate representation, and a deterministic compiler turns
that representation into the database query, so the language model never
writes the query itself. Benchmarks for each module select the model deployed
for that module. The deployed translation model, GPT-5.4-nano, translates 93.7
percent of a hand-built, stratified gold set of 116 queries correctly, where a
translation counts as correct only when both its extracted terms and its
Boolean logic match the reference. The same model builds the strongest keyword
index of any candidate across three scientific keyphrase benchmarks, though on
an extraction task whose scores are low for every method. In the
two ranking stages, recent general-purpose models outperformed the scientific
fine-tuned ones. The leading bi-encoders, Qwen3-Embedding-4B and the deployed
text-embedding-3-small among them, and the leading cross-encoder, Cohere
Rerank 4.0 Pro, score around 82 to 83 on a 0 to 100 ranking-quality scale
(nDCG) on SciRepEval ad-hoc search,
while the fine-tuned SPECTER2 embedders and the biomedical reranker MedCPT
trail the general-purpose leaders. Storing the embeddings as 8-bit integers
costs no measurable ranking quality and halves the memory the index needs.
Together, these component results support the design claim of the thesis,
that structured and semantic retrieval can be combined rather than traded
off. The comparison of the assembled cascade against Google Scholar, PubMed,
and Semantic Scholar is specified in the research design and reported in the
final submission.
\end{abstract}