%=======================================================================
% Analysis and Discussion
% Included from main.tex via \input{...} (or \include{...}).
% Contains the section body only: no \documentclass, no \begin{document}.
% Assumes the IEEEtran class already loaded by main.tex (auto A/B/C and
% 1)/2)/3) numbering for \subsection and \subsubsection).
%
% Revision notes (July 2026 de-duplication pass):
%   - This section now INTERPRETS; the reporting lives in Section IV.
%     Wholesale restatement of Section IV's numbers is removed; each
%     argument keeps only the figures it turns on and cites the table
%     for the rest.
%   - Moved in from Section IV: the reasoning-ablation ceiling/budget
%     reading and the Search saturation mechanism (small re-rank pool).
%   - "modern AI" -> "modern language models" to match RQ1 as reworded
%     in the Introduction.
%   - Short framing lines added under the RQ1 and RQ2 headings so each
%     subsection opens by restating the question it answers.
%   - NEW (item 15): consolidated Limitations subsection added after the
%     RQ3 stub (subsec:limitations). When Section VI is written, its
%     implications/future-work passage should draw on this subsection
%     (the Conclusion rubric criterion asks for exactly that).
%   - Item 14: Appendix~D pointers added at the hallucinated-keyphrase
%     discussion (V-A1) and the filter_distractor discussion (V-A2);
%     the exemplars themselves are being added by the author.
%   - The old header placeholder about \ref{tab:reasoning-effect} was
%     stale: the prose already uses the real label tab:ir-reasoning.
%
% REMAINING PLACEHOLDERS (grep for "TODO"):
%   1. The RQ3 subsection is a stub pending system-level results
%      (analysis framework to be added; see TODO at subsec:rq3).
%   2. The KeyBART/OAGKX fine-tuning claim needs a citation.
%
% NOTE: prose mentions of "the system-level evaluation" are intentionally
% left as plain text, not \ref. Convert to refs later if desired.
%=======================================================================

\section{Analysis and Discussion}
\label{sec:analysis-discussion}

This section interprets the results of Section~\ref{sec:data-results} against the three research questions, taking each in turn. Numbers are restated only where an argument turns on them; the full figures are in the tables cited.

\subsection{RQ1: Structured Graph Retrieval}
\label{subsec:rq1}

RQ1 asks to what extent modern language models can construct the two artefacts that deterministic structured graph retrieval requires: the per-work keyword index on the corpus side, and the intermediate representation on the query side. The two are taken in turn, followed by an overall judgement.

\subsubsection{Keyword Extraction}

GPT-5.4-nano extracts keywords more accurately than every baseline across the three scientific benchmarks, with the single exception of KeyBART on OAGKX. KeyBART's lead there is consistent with its having been fine-tuned on OAGKX, % TODO (item 13): add citation for KeyBART's training data to support this claim.
as the same model falls well below GPT-5.4-nano on both KPBiomed and PubMedAKE (Tables~\ref{tab:kw-kpbiomed} and~\ref{tab:kw-pubmedake}). A lead on the training set that collapses off it is characteristic of a model overfit to one dataset rather than one that extracts well in general.

The extractive quality of GPT-5.4-nano comes at a slight cost. Because it is a language model, GPT-5.4-nano occasionally produces keyphrases absent from the source text despite prompting to suppress this; a representative example, with its source text, is given in Appendix~\ref{app:errors}.
% PLACEHOLDER (appendix D, item 14): error-exemplars appendix to be added by
% the author. Contents: one hallucinated keyphrase alongside the title and
% abstract it was generated from, and one filter_distractor failure with the
% query and the IR the model produced. References left as literal text until
% the appendix exists, per the established convention. The effect is minor: on every dataset its absent (generated) scores fall more than an order of magnitude below its present (extracted) scores, so in practice the model leans heavily toward extraction. The purpose-built extractors, by contrast, score zero on the absent task by construction, since they return only spans that occur in the source.

The reproducibility check (Table~\ref{tab:kw-stability}) adds a qualification the accuracy scores do not capture. With roughly one keyphrase in six changing between runs, the keyword index is not a fixed artefact: two builds of the graph from the same corpus would differ slightly in what they attach to each work. The effect on measured quality is negligible, but the contrast with the baselines, which are reproducible by construction, is a genuine cost of using a language model to construct the corpus-side artefact, and the same cost recurs for the query-side artefact below.

\subsubsection{Intermediate Representation Formation}

The clearest pattern in the IR benchmark (Table~\ref{tab:ir-accuracy}, Figure~\ref{fig:ir-accuracy}) is the gap between the hosted frontier models and every self-hosted open model. The deployed GPT-5.4-nano reaches 0.937 while the best open model, qwen3:8b, reaches only 0.802, and the hosted intervals clear all local ones. The most plausible drivers are model scale and training, though the hosted models' parameter counts are undisclosed, so this remains a hypothesis rather than a measured cause.

Among the open models the generational gap is large. qwen3:4b scores 0.58 higher in absolute terms than its predecessor qwen2.5:3b (Table~\ref{tab:ir-accuracy}), consistent with the broader trend of newer model generations improving on the same task through better training data and methods, and illustrating the advantage a newer variant of the same family can hold. The two smallest and oldest models, phi4-mini and llama3.2:3b, sit near the floor of the task, probably because they have not been trained on tasks resembling this one.

The reasoning ablation (Table~\ref{tab:ir-reasoning}) separates the two hosted models in kind rather than degree. Enabling reasoning moves GPT-5.4-nano by only +0.008 at around 246 reasoning tokens per query, so the model reaches its accuracy in a single pass with no reasoning budget; GPT-5-mini gains +0.190 from a budget nearly four times larger, and only with it reaches a comparable level. The contrast is consistent with GPT-5.4-nano being suited to single-shot use at scale and GPT-5-mini depending on a reasoning budget to compete, though neither model's design intent is documented. Even with reasoning enabled, GPT-5-mini does not meaningfully exceed the deployed model's default configuration, the two scoring within noise of one another, possibly reflecting GPT-5.4-nano being the more recent model of the two.

Within the deployed model, the binding component is the boolean expression rather than the entities (Table~\ref{tab:ir-decomp}). Expression is the harder of the two because the entities are largely surface content lifted directly from the query, whereas the boolean structure is a latent form the model must infer. Even so, most expression errors are near-misses rather than collapses: a graded variant that credits partially correct expressions reaches 0.985, against 0.931 under strict equivalence.

By query type (Table~\ref{tab:ir-stratum}), the errors divide into two kinds. The largest source, \texttt{filter\_distractor}, is a categorisation slip rather than a structural failure: a filter term such as a publication year or citation bound is treated as a searchable entity (a representative failure, with the query and the IR produced, is given in Appendix~\ref{app:errors}), an issue addressable through tighter prompting that defines these cases explicitly.
% PLACEHOLDER (appendix D, item 14): see the note at the Appendix D reference
% in the Keyword Extraction subsubsection above. The remainder fall on the structurally hardest strata, multiple negation, deep nesting, and hard typing, and likely reflect genuine limits of the model's capability rather than a fixable prompt. The perfect scores on the six simpler strata are best read as the absence of observed errors rather than a guarantee of perfect performance, given per-stratum counts of sixteen queries or fewer.

The stability check (Table~\ref{tab:ir-stability}) resolves a reporting question the non-determinism raises. The bootstrap intervals attached to every figure capture sampling uncertainty over the finite gold set; run-to-run variance is a separate quantity, and the check confirms it is negligible, with the model returning identical output on the overwhelming majority of queries at both temperatures. The single-run figures are therefore not undermined by the model's non-determinism, and majority voting over runs adds nothing.

\subsubsection{Overall judgement}

Taken together, these results show that modern language models can construct the artefacts required for deterministic structured graph retrieval, though to differing degrees for the two artefacts involved. On the query side, GPT-5.4-nano constructs the intermediate representation from a natural-language query to a high standard, at a joint accuracy of 0.937, with its residual failures stemming from either genuine model limitations on the structurally hardest queries or loose prompting that is addressable. On the corpus side, the same model produces the strongest benchmarked keyword index of any candidate, but only in relative terms, since keyphrase extraction is a task where absolute scores are constrained by well-documented ceilings in the F1@5 metric, and GPT-5.4-nano's results are competitive with published extraction performance on comparable scientific datasets. Both artefacts also carry the cost of being produced by a language model rather than a deterministic process: the keyword stage occasionally generates rather than faithfully extracting, and neither stage is fully reproducible across runs.

Because both artefacts feed a deterministic matcher, the quality of the structured stage as a whole is bounded entirely by how well these two can be constructed, and on this evidence that bound is high on the query side and competitive but imperfect on the corpus side. What neither component score establishes on its own is how the two behave in combination over the full corpus, which is the question the system-level evaluation (Section~\ref{subsec:rq3}) takes up, where the composed system is compared against existing tools. It is likely, too, that models better suited to both tasks will continue to emerge.

\subsection{RQ2: Ranked retrieval mechanism}
\label{subsec:rq2}

RQ2 asks how well current bi-encoders and cross-encoders perform semantic ad-hoc search on scientific retrieval benchmarks. Each stage is analysed against its results in Section~\ref{sec:data-results}, then the two are judged together.

\subsubsection{Bi-encoders}

The immediate observation is that the general-purpose embedding models outperform the fine-tuned models on average, with both SPECTER2 variants among the bottom three (Table~\ref{tab:bienc-results}). This is consistent with the fine-tuned models having been trained for one task type rather than for retrieval in general. SPECTER2-adhoc-query, tuned for the query side of ad-hoc search, tops the Search task clear of the field, yet records the worst NFCorpus score of any model and the third-lowest average. Its Search lead is therefore close to in-domain performance rather than evidence of general retrieval quality, and it does not carry over to the other tasks. SPECTER2-base sits at the bottom on average for a different reason, in that it is not tuned for ad-hoc search at all. The general-purpose models, trained across many domains, hold up across all three tasks, which is what places them above both fine-tuned variants.

Unlike IR extraction, the hosted model is not the winner here. text-embedding-3-small sits inside the five-way statistical tie at the top but below three open models on point estimate (Table~\ref{tab:bienc-results}), and Qwen3-Embedding-4B's lead is plausible rather than established, with only marginal overlap between the lower tail of its interval and the upper tails of the rest. Across tasks, the selected model is level with the leaders on Search and close on NFCorpus, the two broader benchmarks; its largest gap is on TREC-CoVID, where its point estimate sits below the Qwen models with only marginal interval overlap. This may suggest that it weakens on more specific benchmarks as queries tighten, though TREC-CoVID is the only such benchmark here, so the reading is suggestive rather than established.

EmbeddingGemma-300M is the most deployment-efficient of the strong models. It matches the leading cluster on nDCG while being the smallest competitive model on both axes that drive cost (Table~\ref{tab:bienc-deploy}): it is around thirteen times smaller than Qwen3-Embedding-4B and half the size of Qwen3-Embedding-0.6B by parameter count, and it carries the smallest embedding dimension of any competitive model at 768. The dimension is the more consequential of the two, since in-memory footprint scales with it: at int8 over the 85 million works, 768 dimensions require around 65 GB against around 131 GB for text-embedding-3-small at 1536. A lighter model also encodes more cheaply. EmbeddingGemma therefore reaches the leading cluster's accuracy at a fraction of its memory and compute.

There is nonetheless a real advantage to a hosted model such as text-embedding-3-small, in that it provides managed infrastructure with no upfront compute cost, whereas the self-hosted models require provisioning hardware to encode all 85 million works directly. The tradeoff is two-sided: the hosted model is a paid API with undisclosed parameters, carrying recurring cost and no control over the model, while self-hosting carries a one-time compute cost and full control. This tradeoff is what the selection turns on, and it is taken up in the overall judgement.

\subsubsection{Cross-encoders}

For the cross-encoders there is no significant difference between cloud-hosted and self-hosted models. Cohere Rerank 4.0 Pro leads on point estimate, but its interval overlaps those of the next three models, so the top four form a genuine statistical tie (Table~\ref{tab:xenc-results}), and within it the strongest self-hostable model, Qwen3-Reranker-4B, is tied with the hosted leader. Unlike the bi-encoder, where the hosted model sat mid-cluster, here a self-hostable model reaches the top of the benchmark.

The models that lead are particular newer families rather than the larger ones. The Cohere 4.0 and Qwen3 rerankers occupy the top of the table across sizes, and parameter count does not track performance: Qwen3-Reranker-0.6B at 82.43 outscores much larger but older models such as bge-reranker-large at 76.00. The differentiator is therefore the model family and its training rather than raw scale. The one domain-specific model, MedCPT, sits mid-pack below the strong general models (Table~\ref{tab:xenc-results}), even though the corpus is biomedical and MedCPT is the biomedical-trained reranker. For re-ranking, then, domain specialisation offered no advantage, and the general models outperformed the specialist on its own domain, the opposite of the in-domain effect seen for the fine-tuned bi-encoders.

The per-task pattern also shows where re-ranking can and cannot matter. Nearly all of the seventeen-point spread between rerankers comes from NFCorpus, whereas on Search every model scores close to 75, because the small candidate pool the stage re-ranks leaves little room to reorder, so the task barely separates the models. The same property applies inside the cascade, where the reranker sees only the roughly 100 candidates the earlier stages pass forward: the stage's contribution is bounded by the pool it is given, which is part of what the system-level evaluation must establish.

Unlike the bi-encoder, model size carries no real advantage here. The stage re-ranks only around 100 candidates at a time, so its cost is query-time latency rather than stored memory, and at that volume the difference between a small and a large model is a matter of milliseconds. The contrast with the bi-encoder is structural: the bi-encoder embeds and stores a vector for all 85 million works, so embedding dimension drives a permanent memory cost, whereas the reranker stores nothing and runs only over the small candidate pool at query time. Size was therefore decisive for the bi-encoder and is close to irrelevant here, which means the most accurate reranker can be chosen without the deployment penalty that constrained the embedding choice.

The infrastructure tradeoff still holds but applies on a different axis. Because re-ranking is bound by per-query latency rather than a fixed memory footprint, its cost scales with query volume rather than corpus size. As the user base grows, the pressure falls on throughput and concurrent query handling, which is the opposite axis to the bi-encoder's fixed cost of storing 85 million vectors. Scaling the reranker is therefore a matter of serving capacity rather than storage.

\subsubsection{Overall judgement}

Taken together, these results show that current bi-encoders and cross-encoders perform well on the scientific ad-hoc search benchmarks, with the leading models in both stages clustered near the top of each table at averages around 82 to 83 nDCG. Both semantic stages of the cascade can therefore be equipped with capable models, and the retrieve-then-rerank pattern they implement is well-supported by what is currently available.

Across both components, the models that lead are particular recent families rather than the domain or task-specific ones. The Qwen3 and Cohere 4.0 families and the general-purpose embedders sit at the top, while the specialist models do not lead their tasks, though the effect is not uniform: in-domain training helped the bi-encoder on its in-domain task, where SPECTER2-adhoc-query topped Search, yet did not help the reranker on its in-domain corpus, where the biomedical MedCPT sat mid-pack on biomedical text. Most of the gaps between the leaders are within confidence intervals, so the leading models are better read as a strong cluster than as a clear ranking.

In neither stage does the paid hosted model significantly outperform the open ones on accuracy. For the bi-encoder, text-embedding-3-small sits within the leading statistical tie but below three open models on point estimate; for the cross-encoder, Cohere Rerank 4.0 Pro leads on point estimate yet ties the open Qwen3-Reranker-4B within intervals. In both stages, accuracy alone does not separate the hosted option from the best open one, so the selection turns on deployment characteristics rather than retrieval quality.

On that basis both stages were assigned to the hosted models for v1, for managed infrastructure with no upfront compute: the self-hosted alternatives would require provisioning hardware to encode all 85 million works, whereas the hosted APIs carry a recurring cost but no build-out. This is an engineering choice, not a quality one, and it is reversible. Because the tied open models, EmbeddingGemma-300M for the bi-encoder and Qwen3-Reranker-4B for the cross-encoder, are accuracy-viable in their own right, the system can move to self-hosting without loss of retrieval quality should the cost structure or scale make that preferable.

A limit applies to all of the above. These are full-corpus benchmark scores, which stand as a proxy for the role each model plays inside the cascade, where it ranks the structured candidate pool the first stage returns rather than the full corpus. Whether the benchmark ranking holds over that candidate pool is a property of the integrated system, and is taken up in the system-level evaluation (Section~\ref{subsec:rq3}), where the cascade is run end to end over real queries.

\subsection{RQ3: Comparison of System Against Others}
\label{subsec:rq3}

% TODO (item 16): expand this stub into an analysis framework ahead of the
% results: the comparison dimensions (constraint satisfaction, semantic
% recall, auditability of the returned set, coverage) and the hypotheses the
% RQ1/RQ2 results generate for how the cascade should differ from a
% semantic-only LLM and from Google Scholar.
This section answers RQ3 by comparing the assembled cascade against a retrieval-augmented LLM and Google Scholar over a set of realistic queries. RQ1 and RQ2 bound the structured and ranked stages in isolation; the comparison here explains how the composed system's results differ from those of a semantic-only LLM and a text-plus-citation engine, grounded in the architectural differences those stages established. It is qualitative, characterising per query what each system surfaces and misses, supported by light relevance judgments. Results are reported in the final submission.

\subsection{Limitations}
\label{subsec:limitations}

The analysis above notes several limits where they arise; this subsection gathers them and adds those that span components rather than belonging to any one.

The strongest caveat on RQ1 concerns the gold set. It was curated and annotated by a single author, so the reference IRs embody one reader's judgement of each query's intended structure, and no measure of inter-annotator agreement was possible. Queries written by the system's designer also risk matching the input the system was built to expect, a construct-validity concern that the stratified design and the sourcing of queries from SciRepEval and systematic-review structures reduce but do not remove. At 116 queries the per-stratum counts are small, which Section~\ref{sec:data-results} already treats by reading failure counts rather than rates.

On RQ2, the component scores stand in for behaviour inside the cascade, the proxy limit stated in the overall judgement above. Two further features of the benchmark deserve note. TREC-CoVID contributes only fifty queries, so its intervals are wide, and most of the between-model separation among the rerankers comes from a single collection, NFCorpus, with Search saturated and TREC-CoVID noisy; the reranker ranking therefore rests more heavily on one dataset than the three-task average suggests. The evaluation as a whole is also confined to the biomedical deployment, so the schema's claimed generality beyond that domain is a design property rather than a measured one.

A second family of limits comes from the substrate. Section~\ref{sec:lr-limitations} documented that OpenAlex retains references to deleted works and that over sixty percent of its journal articles lack complete institutional metadata. % VERIFY: intended target is the substrate-flaws discussion (II-E1); confirm the label resolves there once Section II is revised. TODO (item 13): optionally repeat the Culbert et al. and Zhang et al. citations here.
The second gap bears directly on the structured stage's headline capability. An institution-anchored query is answered through the affiliation edges the graph contains, so a work whose affiliation is missing from OpenAlex is silently excluded from an AND over that institution, and the candidate pool undercounts precisely where the typed graph is meant to be strongest. The architecture makes such gaps inspectable, in that works lacking affiliations can themselves be queried, but the retrieval path does not currently surface the exclusion to the user, and its size is not quantified here. The published figures are also corpus-wide, and how the missing metadata distributes over the biomedical subset is not known.

A third limit is structural. The boolean gate means that, for topical entities, stage-one recall is bounded by the offline keyword index: a work whose extracted keywords fail to capture a concept can never enter the candidate pool, and no later stage can recover it. Exact-match terms and non-keyword entities such as authors, institutions, and venues follow separate match paths and are not subject to this bound. The component evaluation measures the index's agreement with reference keyphrases, not the candidate-pool recall that follows from it, so the practical cost of index imperfection is not directly measured by the benchmarks of Section~\ref{sec:data-results} and falls to the system-level evaluation. The related cost of building both artefacts with a language model, that neither is fully reproducible across runs, is discussed under RQ1 (Section~\ref{subsec:rq1}).

Finally, the deployed configuration assigns all three learned components to hosted models served through paid APIs with undisclosed parameters. The benchmark results are therefore tied to those endpoints as evaluated, and provider-side changes in model version or behaviour cannot be ruled out over the system's life. The mitigation is the one the overall judgement of RQ2 identified: for the two semantic stages, accuracy-viable open alternatives exist, so the dependence is a reversible engineering choice rather than an architectural commitment. For IR extraction the gap to the best open model is wide, and that dependence is correspondingly harder to unwind.