% =====================================================================
% Literature review, Section II-B: recent research advances.
% Structure: every paragraph opens with the point it exists to make,
% evidence and grouped citations follow, and the paragraph closes by
% tying the point to this thesis's design.
% =====================================================================

\subsection{Recent research advances}
\label{sec:recent-advances}

Five threads of recent work bear on the design, and this thesis takes a
different stance toward each: dense retrieval and the cross-encoder
re-ranking reviewed alongside it are adopted as the cascade's two
ranking stages (Section~\ref{sec:lr-dense}); graph augmentation is
treated as motivation for the typed graph rather than as a baseline to
copy (Section~\ref{sec:lr-graph}); structured-query translation is the
design pursued for the first stage (Section~\ref{sec:lr-nl2q}); query
reformulation is set aside (Section~\ref{sec:lr-rag}); and keyphrase
extraction supplies the corpus-side index that the structured stage
matches against (Section~\ref{sec:lr-keyphrase}).  The grounds for each
stance open the corresponding subsection.

\subsubsection{Scholarly embeddings, dense retrieval, and re-ranking for scientific text}
\label{sec:lr-dense}

Dense retrieval and cross-encoder re-ranking are retained as this
thesis's two ranking stages rather than as its retrieval mechanism.
The strongest single-pass models still miss roughly a quarter of
relevant papers on realistic scientific queries, which rules dense
retrieval out as the primary stage, while the accuracy that joint
scoring recovers earns the cross-encoder the final one.

Three lines of progress have improved dense retrieval without removing
the failure that matters most here, which is unreliable behaviour on
the entity-heavy and lexically precise queries that a structured stage
answers by construction.  Scaling the training distribution, from DPR
\cite{karpukhin2020dpr} through unsupervised contrastive pre-training
\cite{izacard2022contriever} to weakly supervised pairs at web scale
\cite{wang2022e5,li2023gte,xiao2023cpack,chen2024bgem3}, lifted average
quality on the BEIR suite \cite{thakur2021beir} while leaving exactly
that weakness in place.  Late interaction
\cite{khattab2020colbert,santhanam2022colbertv2} and learned sparse
retrieval \cite{formal2021splade,formal2022splade} showed instead that
preserving finer-grained interaction between query and document terms
buys accuracy at modest cost, and ColBERTv2 does so while shrinking a
late-interaction index by a factor of 6 to 10
\cite{santhanam2022colbertv2}.  The cross-encoder stage below pushes
the same trade to its limit.  The scholarly-specific line pursued
citation supervision
\cite{beltagy2019scibert,cohan2020specter,ostendorff2022scincl}, and
its lesson for this thesis is negative.  The SciRepEval benchmark
\cite{singh2023scirepeval} shows that proximity-trained representations
transfer poorly to classification and ad-hoc search, with the
adapter-based SPECTER2 only a few MAP points ahead of a BM25 baseline
on multi-domain citation recommendation.  Model selection for the dense
stage therefore cannot assume that scholarly fine-tuning wins, which is
why the research design benchmarks general-purpose and domain-tuned
encoders side by side.

Cross-encoder re-ranking recovers the accuracy that separately computed
representations give up, and this thesis retains it as the final stage
of the cascade.  Nogueira and Cho's BERT passage re-ranker, later
termed monoBERT \cite{nogueira2019monobert}, established the design by
reading the query and the document together in one input, so that every
query term attends to every document term, and monoT5 extended the same
idea to sequence-to-sequence scoring \cite{nogueira2020monot5}.  Joint
reading is what late interaction only approximates and what
single-vector encoders cannot express at all, and it carries a defined
price.  Every query and document pair requires a fresh forward pass, so
nothing can be precomputed and the model is affordable only over a
small candidate set.  Pairing a cheap high-recall first stage with an
expensive cross-encoder over its top candidates is the established
multi-stage ranking pattern \cite{nogueira2019multistage}, and the two
semantic stages of this thesis instantiate it, with a bi-encoder
ranking the structured candidate pool and a cross-encoder re-ranking
the shortlist it returns.  The research design selects both models by
benchmark rather than by reputation.

The size of the gap that remains is established by LitSearch
\cite{ajith2024litsearch}, a benchmark of 597 literature-search
questions either written by the papers' own authors or rewritten from
inline citations, and it is the source of the quarter-missed figure
this subsection's stance rests on.  The best open model tested,
GritLM-7B, reaches Recall@5 of 74.8 on specific questions, so roughly
one relevant paper in four is absent from its top results, and deployed
commercial tools reach only 17.5 on the same author-written questions.
The difficulty is scholarly search itself rather than retrieval at
large, since the same model's nDCG@10, which scores how highly the
relevant papers sit within the top ten results, falls from 70.3 on
general-domain Natural Questions to 44.1 on broad LitSearch queries.
Two of the benchmark's failure modes matter directly here.  Queries
whose wording shares little with the target paper's title and abstract
are systematically harder, because retrievers exploit lexical overlap,
and broad questions are harder than specific ones.  Both are weaknesses
of matching in a single learned space, and neither touches the further
class of query, over typed relations and multiple citation hops, that
no single-vector representation can express at all.  Related benchmarks
reinforce the reading that scholarly-specific embeddings are not the
remedy, with DORIS-MAE \cite{wang2023dorismae} finding general-purpose
models modestly ahead on multi-aspect queries and CSFCube
\cite{mysore2021csfcube} finding SPECTER unstable across facets.  These
failures are what the structured graph stage of this thesis exists to
remove, and they are the case for adding graph traversal alongside text
and embedding lookup.

\subsubsection{Graph-augmented retrieval}
\label{sec:lr-graph}

Graph-augmented methods supply the conceptual motivation for ingesting
OpenAlex into a typed knowledge graph but offer no benchmarked baseline
to copy on realistic scientific queries, so this thesis treats the
literature as identifying the gap rather than filling it.  Scholarly
corpora are already graphs, in that papers cite papers, authors
co-author, venues publish, and funders grant, and three sub-threads
have been proposed to exploit that structure, namely knowledge-graph
embeddings, graph neural networks, and Personalized PageRank.  Each has
been evaluated almost entirely off scholarly search.

Knowledge-graph embeddings offer no evidence of retrieval utility,
because the family is validated on link prediction rather than search.
The canonical models TransE, ComplEx, and RotatE
\cite{bordes2013transe,trouillon2016complex,sun2019rotate} are
benchmarked almost exclusively on FB15k and WN18 variants, graphs
unlike scholarly ones in scale, edge directionality, and node
heterogeneity.  The one application at scholarly scale makes the gap
explicit, in that F\"arber and Ao \cite{farber2022makgenhanced} report
near-perfect link prediction over MAKG (ComplEx MRR 0.958) yet evaluate
no literature-search task at all.  Accuracy at predicting missing edges
is not evidence that the embeddings can rank papers against a
natural-language query.

Graph neural networks fail the same relevance test for a different
reason, in that their evaluation regimes assume graphs and tasks that
scholarly search does not have.  GraphSAGE \cite{hamilton2017graphsage}
and GAT \cite{velickovic2018gat} are validated on node classification
over citation networks orders of magnitude smaller than OpenAlex, and
LightGCN \cite{he2020lightgcn} assumes the recurring users and
symmetric interactions of recommendation, neither of which a
directional, time-stamped citation graph with an extreme long tail
provides.  Scholarly-specific neural work has moved past those
assumptions \cite{bhagavatula2018citeomatic,shen2024tgntrec} but has
produced no benchmark comparable in realism to LitSearch, so it
constrains nothing in this design.

Personalized PageRank is the one graph method whose mechanics match the
retrieval problem, and even it lacks a scholarly-search evaluation.
Haveliwala's topic-sensitive formulation \cite{haveliwala2002tspr}
conditions the walk on the query through the teleport vector, so scores
reflect query-specific relevance rather than generic prominence, and
the propagation aggregates influence across citation hops that no
single-vector representation reaches.  Chen et al.\ \cite{chen2007gems}
use exactly this to surface pivotal but under-cited physics papers.
The scholarly-search literature between 2020 and 2026 nevertheless
contains no rigorous combination of PPR with modern dense retrieval
evaluated on scientific information retrieval.  The closest analogue, HippoRAG
\cite{jimenezgutierrez2024hipporag}, seeds PPR from query entities over
an LLM-extracted graph but evaluates on Wikipedia-derived multi-hop
question answering rather than scientific corpora.

The modern graph-RAG variants confirm the pattern rather than closing
the gap.  GraphRAG \cite{edge2024graphrag} and LightRAG
\cite{guo2024lightrag} evaluate community-summary question answering
over podcasts, news, and textbooks, and the strongest scholarly RAG
system, OpenScholar \cite{asai2024openscholar}, abandons the citation
graph entirely in favour of a text retriever with iterative
self-feedback.  Graph structure is widely asserted to help scholarly
search and almost never benchmarked for it against modern dense
baselines.

This thesis takes that conjunction of strong conceptual fit and thin
empirical record as motivation rather than as a baseline to copy.  It
ingests OpenAlex into a typed knowledge graph and exposes the structure
through traversal queries directly, treating the KGE, GNN, and PPR
literature as work that identifies the gap rather than work that fills
it.  The generative components of GraphRAG and its successors are out
of scope, since the target here is first-stage retrieval returning
ranked paper lists rather than synthesised answers.  How a
natural-language question becomes a structured query against an
OpenAlex-derived schema is the question Section~\ref{sec:lr-nl2q} takes
up.

\subsubsection{Natural-language-to-structured-query systems for scholarly knowledge graphs}
\label{sec:lr-nl2q}

Structured-query translation is the design this thesis pursues,
instantiated through an intermediate JSON representation that decouples
LLM-based intent extraction from deterministic query compilation, and
this third position avoids the structural weaknesses of both the
existing scholarly benchmarks and end-to-end LLM tool calling.

The translation technology itself is mature, and its trajectory in the
SQL setting defines what a credible evaluation must contain.  Spider
\cite{yu2018spider} and BIRD \cite{li2023bird} established cross-domain
evaluation with disjoint training and test schemas, and decomposition
methods such as DIN-SQL \cite{pourreza2023dinsql} brought models close
to human performance on them.  Spider 2.0 \cite{lei2024spider2} then
showed that ceiling to be an artefact of small schemas, dropping GPT-4o
to roughly 10\% success once enterprise-scale schemas appear.  The
lesson carried into this thesis is that translation accuracy means
little unless a realistic schema is supplied at inference, which is how
its own gold set is constructed.

The scholarly-specific benchmarks are too weak to reuse, which is why
this thesis builds its own.  SpCQL \cite{guo2022spcql} pairs 10,000
Chinese questions with Cypher over an open-domain graph but leaves the
schema unspecified at inference.  DBLP-QuAD \cite{banerjee2023dblpquad}
and SciQA \cite{auer2023sciqa} generate their pairs from hand-written
templates, and the spread reported on SciQA in F1, the harmonic mean of
precision and recall, runs from under 26\% for zero-shot models to over
97\% after fine-tuning, which is as much evidence of template
saturation as of progress.  Both SPARQL benchmarks further assume gold
entity linking that no deployed system receives, and QALD
\cite{usbeck2023qald10} remains general-domain.  A benchmark that
rewards template memorisation under gold linking cannot certify the
translator this pipeline needs.

Tool-calling agents sit at the opposite pole, trading grammar-level
correctness guarantees for flexibility.  ReAct \cite{yao2023react}
interleaves free-text reasoning with executable actions, Toolformer
\cite{schick2023toolformer} shows a model can learn when to call tools,
and function-calling APIs have since made the pattern the default
substrate for agentic systems.  Query construction moves into the
prompt and the tool schema, so nothing guarantees that the emitted
query is valid or optimal, and direct comparisons between fine-tuned
parsers and tool-using agents on scholarly KGQA remain scarce.

This thesis's design occupies a third position between those poles.
Rather than fine-tuning a grammar-constrained parser end-to-end (the
SpCQL, DBLP-QuAD, and SciQA tradition) or letting an LLM emit Cypher
directly through function calling (the ReAct and Toolformer tradition),
it introduces an intermediate JSON representation that names typed
entities and the boolean expression over them.  The LLM populates the
structure from natural language, and a deterministic compiler
translates it into Cypher against a fixed schema derived from OpenAlex.
The split inherits compositional flexibility from the function-calling
line, since the LLM never needs to know the graph language, and keeps
the correctness guarantees of the parser line, since query construction
never leaves the compiler.  It also avoids the empirical weaknesses
above, in that there are no templates to memorise, gold entity linking
is not assumed, and the schema is supplied at inference.  Whether this
design generalises to the structural queries reformulation cannot reach
is the question Section~\ref{sec:lr-rag} takes up.

\subsubsection{LLM query reformulation for scholarly search}
\label{sec:lr-rag}

This thesis sets LLM query reformulation aside, both because it
degrades the strong retrievers it would need to combine with and
because it cannot in principle express the structural queries a typed
graph makes available.  The methods leave the retriever unchanged and
rewrite the query instead, whether by encoding a hypothetical answer
document, appending an LLM pseudo-document, or generating expansion
text in place of classical relevance feedback
\cite{gao2023hyde,wang2023query2doc,jagerman2023qexp,mackie2023grf}.

The decisive evidence is Weller et al.\ \cite{weller2024whenfail}, who
ran eleven expansion methods across twelve datasets and twenty-four
retrievers and found expansion gains negatively correlated with
retriever strength.  Expansion helps weak retrievers and routinely
hurts strong ones such as ColBERTv2 and SPLADE, with the generated text
diluting the original relevance signal.  Two implications follow for
this design.  A pipeline built on strong dense retrieval over a typed
OpenAlex-backed graph sits exactly where expansion is documented as
negligible or negative, and no expansion, however semantically rich,
can express a structural query such as \emph{papers by authors at
institution X that cite work from lab Y}.  The structured-parsing path
of Section~\ref{sec:lr-nl2q} is therefore the right tool for this
thesis and reformulation is not.  The adjacent scholarly-RAG systems
that would consume such pipelines target end-to-end answer generation
rather than first-stage retrieval, and
Section~\ref{subsec:llm-agentic} surveys them as users of this system's
output rather than competitors to it.

\subsubsection{Keyphrase extraction for scholarly text}
\label{sec:lr-keyphrase}

The corpus-side artefact of the structured stage is a per-work keyword
index, so the other half of RQ1 sits in the keyphrase-extraction
literature, which this thesis draws on for baselines and evaluation
conventions rather than for a deployable extractor.  Meng et al.\
\cite{meng2017deepkp} reframed the task as sequence-to-sequence
generation and introduced the now-standard split between present
keyphrases, which occur verbatim in the source text, and absent
keyphrases, which must be generated.  The split matters here because an
index built by a language model can hallucinate terms the source does
not contain, a failure purely extractive methods cannot commit, so the
present score isolates the faithfulness a deterministic matcher depends
on.  Three families supply the baselines this thesis benchmarks.
Statistical and graph-based extractors score candidate spans without
training and can recover only present keyphrases, running from TF-IDF
term weighting \cite{salton1988termweighting} through the
graph-centrality methods TextRank \cite{mihalcea2004textrank} and
MultipartiteRank \cite{boudin2018multipartiterank} to YAKE's
corpus-independent features \cite{campos2020yake}.  Neural extractors
such as KeyBERT \cite{grootendorst2020keybert} and the keyphrase-aware
pre-trained KBIR \cite{kulkarni2022kbir} improve span ranking but
remain extractive.  Generative models can also produce absent
keyphrases, with KeyBART \cite{kulkarni2022kbir} the representative
fine-tuned generator and a prompted general-purpose language model the
newest member of the family.  Evaluation is by F1 at a small cutoff
against author-assigned gold keyphrases, usually with stemming, and its
absolute ceiling is low because author keyphrase sets are short,
subjective, and only partly recoverable from the title and abstract
alone, so even strong systems score in the tens, as the biomedical
PubMedAKE benchmark shows directly \cite{sheng2022pubmedake}.  The
datasets this thesis evaluates on follow that convention at scale, with
OAGKX deriving some 23 million records from the Open Academic Graph
\cite{cano2020oagkx} and KPBiomed pairing more than five million PubMed
abstracts with their author keyphrases \cite{houbre2022kpbiomed},
alongside PubMedAKE over the same source \cite{sheng2022pubmedake}.
What agreement with authors does not measure is the retrieval the index
exists to serve, and the closest result is Boudin et al.\
\cite{boudin2020kpretrieval}, who show that appending generated
keyphrases to documents improves scientific ad-hoc retrieval.  That gap
between agreement and retrieval recurs later as a limitation of this
thesis's own component evaluation.