%=============================================================================
% CASE STUDY 2 - COMP4318 (Machine Learning and Data Mining)
%-----------------------------------------------------------------------------
% Wired into the appendix block in main.tex directly after the COMP3308 case
% study, so it letters itself as Appendix F automatically.
%
% \ref keys below are wired to the real labels in the report. Keys marked
% [shared] resolve to the same label used by case_study_comp3308.tex:
%   subsec:system-architecture   [shared]  Sec III-B  System Architecture
%   sec:lr-dense                 [shared]  Sec II-B1  Dense retrieval
%   sec:lr-graph                 [shared]  Sec II-B2  Graph-augmented retrieval
%   subsec:rq1 / subsec:rq2      [shared]  Sec V-A / V-B  RQ1 / RQ2 analysis
%   sec:data-results             [shared]  Sec IV     Data and Results
%   sec:analysis-discussion      [shared]  Sec V      Analysis and Discussion
%   app:prompts                  [shared]  App C      Prompts and Tool Schemas
%   tab:ir-stratum               [shared]  Table X    tab:ir-reasoning  Table VIII
%   tab:kw-stability             [shared]  Table VII  tab:ir-stability  Table XI
%   subsec:traditional-search    Sec II-C2  Traditional scholarly search engines
%   subsubsec:ingestion-storage  Sec III-C2 Ingestion and Storage
%   subsec:evaluation-validation Sec III-D  Evaluation and Validation Design
%   subsubsec:keyword-generation Sec III-D1 Keyword Generation (validation)
%   subsubsec:embedding-quantisation Sec III-D5 Embedding Quantisation (validation)
%   sec:results-keywords         Sec IV-A   Keyword Generation (results; label added there)
%   tab:bienc-deploy             Table XII  tab:ir-accuracy  Table XIV
%=============================================================================

\section{Case Study 2: COMP4318 (Machine Learning and Data Mining)}
\label{app:comp4318}

This case study evidences the learning outcomes of COMP4318, Machine
Learning and Data Mining, against the work documented in this report. In
the format agreed with the academic supervisor, an outcome the placement
exercised is evidenced by cross-reference to the body of the report, and
new material is written in place only where the placement did not itself
exercise the outcome, which here applies to the reinforcement-learning
component of LO1 alone. The mapping follows the scoping recorded in the
progress report.

\subsection{LO1: Classification, regression, clustering and reinforcement learning}
\textit{Understand the basic principles, strengths, weaknesses and
applicability of machine learning algorithms for solving classification,
regression, clustering and reinforcement learning tasks.}

This outcome asks for understanding and analysis rather than
implementation, which LO2 covers, so the evidence is drawn both from the
system's own learned components and from the report's critical treatment
of machine learning deployed by others. For classification,
Section~\ref{subsec:traditional-search} analyses scite's citation-intent classifier,
a fine-tuned SciBERT model labelling citation statements as supporting,
contrasting or mentioning, against the third-party audit that measured
F-scores between 0 and 0.58 where the vendor claimed over 80 percent
precision, a concrete lesson in the gap between reported and audited
classifier performance. The system contains a classification decision of
its own, since IR extraction assigns each extracted span one of four
entity types and must separate searchable entities from filters. The
error analysis of Section~\ref{subsec:rq1} traces the deployed model's
largest failure mode, the \texttt{filter\_distractor} stratum of
Table~\ref{tab:ir-stratum}, to precisely that boundary, a year or citation
bound miscategorised as a searchable entity.

For regression, the cascade's ranking stages are pointwise learned
scorers. The cross-encoder regresses a single continuous relevance score
for each query--document pair, which alone determines the final ordering
(Section~\ref{subsec:system-architecture}), and the bi-encoder scores candidates by
cosine similarity between learned representations. This is where the
progress report's planned regression-based relevance scoring across graph
and embedding signals landed, with the hybrid resolving into a cascade in
which each learned scorer ranks within the pool the previous stage passes
forward. The strengths-and-weaknesses analysis of the two scorer designs
runs through Sections~\ref{sec:lr-dense} and~\ref{subsec:rq2}, where
independent encoding permits precomputation over 85 million works but
cannot model query--document interaction, while joint scoring recovers
that interaction at a per-pair cost that confines it to the final 100
candidates. The dense stage is also an exact nearest-neighbour search in
a learned space, and it exhibits the classic instance-based trade of no
per-query training against a memory footprint that scales with corpus
size and embedding dimension, the pressure that motivates the int8
storage quantisation (Section~\ref{subsubsec:ingestion-storage},
Table~\ref{tab:bienc-deploy}).

For clustering, the applicability analysis runs in the negative
direction. Section~\ref{sec:lr-graph} examines GraphRAG, which clusters
an LLM-extracted graph with Leiden community detection and answers
sensemaking questions from community summaries, and judges the approach
out of scope because the target here is first-stage retrieval returning
ranked paper lists rather than synthesised answers, a judgement about
what clustering-based aggregation is and is not the right tool for.
Clustering also appears inside a benchmarked baseline, MultipartiteRank,
which groups candidate keyphrases into topics before ranking them, and
its behaviour against the non-clustering extractors is measured across
all three keyphrase datasets.

The placement presented no reinforcement-learning task, and none is
claimed. Its learning problems, offline keyphrase extraction over a fixed
corpus and single-shot scoring of a query against candidates, contain no
sequential decision process, no environment transitioning under the
system's actions, and no reward signal to optimise against, so an RL
formulation was not warranted. Recognising when the paradigm does not
apply is itself the applicability judgement this outcome names. The
closest empirical contact is indirect. The hosted extraction models
belong to families whose providers attribute their reasoning behaviour to
reinforcement-learning-based post-training, and the reasoning ablation of
Table~\ref{tab:ir-reasoning} measures what that capability contributes at
inference on this task, nothing for the deployed model at +0.008 and a
decisive +0.190 for GPT-5-mini. The attribution to the training method is
the vendor's, since the pipelines are proprietary.

\subsection{LO2: Designing, implementing and evaluating machine learning algorithms}
\textit{Have obtained practical experience in designing, implementing and
evaluating machine learning algorithms.}

The implementation experience spans the ML-facing parts of both
pipelines, built from scratch for this project. On the ingestion side sit
the keyword-generation and embedding-generation modules, scaled from the
one-million-node test ingest recorded in the progress report to the full
85-million-work subset (Section~\ref{subsubsec:ingestion-storage}), and on
the evaluation side sit the benchmark harnesses for keyword extraction
and IR extraction, including the hand-built 116-query stratified gold
set. The one disclosed exception is the SciRepEval evaluation harness,
which was adapted from its official release and extended so that the
non-SPECTER candidates, hosted and self-hosted alike, could be wired into
the same protocol. The evaluation experience is ML-specific in its
choices. Metrics are matched to task, nDCG for ranking and F1@5 with
stemming for keyphrases, with the present/absent split designed to
separate faithful extraction from generation
(Section~\ref{subsubsec:keyword-generation}). Candidates exceeding the
8\,GB evaluation GPU run under post-training weight-only quantisation
with the precision annotated per model
(Section~\ref{subsec:evaluation-validation}), an ablation isolates the
accuracy cost of int8 storage quantisation
(Section~\ref{subsubsec:embedding-quantisation}), and run-to-run
stability is quantified for the stochastic components
(Tables~\ref{tab:kw-stability} and~\ref{tab:ir-stability}). The
evaluations also changed the system rather than merely describing it. The
keyword extractor deployed at the progress report, GPT-5.1 mini, was
superseded by GPT-5.4-nano when the benchmarking of
Section~\ref{sec:results-keywords} selected it, the
substitution-without-redesign the architecture was built to permit.

\subsection{LO3: Machine learning software and libraries}
\textit{Have gained practical experience in using machine learning
software and libraries.}

The progress report recorded that the broader framework use envisioned in
the proposal was not required under the revised scope, and that scoping
holds. The placement's library experience lies not in training models but
in selecting, quantising, serving and evaluating them, which is the form
the skill takes in deployed machine-learning work. The stack in use spans
four serving regimes driven through one evaluation protocol. Self-hosted
embedders and rerankers run on PyTorch through the HuggingFace
Transformers and Sentence-Transformers libraries, with weight-only
quantisation applied to the candidates exceeding the evaluation GPU
(Section~\ref{subsec:evaluation-validation}). The local IR-extraction
candidates are served through Ollama at 4-bit quantisation
(Table~\ref{tab:ir-accuracy}). The hosted models are driven through the
Azure OpenAI and Cohere SDKs, including forced tool calls against a typed
schema (Appendix~\ref{app:prompts}), and the keyphrase baselines run
through their library implementations, KeyBERT, YAKE and the pke toolkit.
% VERIFY: confirm pke is the implementation used for TextRank,
% MultipartiteRank and TF-IDF; adjust the library names to the actual
% imports before submission.
Inside the database, Memgraph's MAGE library supplies the
cosine-similarity computation over the stored int8 vectors
(Section~\ref{subsec:system-architecture}), and the bootstrap confidence intervals
reported throughout Section~\ref{sec:data-results} are computed with NumPy.
% VERIFY: confirm NumPy (vs SciPy or pandas) for the bootstrap
% implementation.

\subsection{LO4: Presenting and interpreting data in verbal and written form}
\textit{Present and interpret data and information in verbal and written
form.}

The written evidence comprises the progress report and this report, in
which Sections~\ref{sec:data-results} and~\ref{sec:analysis-discussion} present the
benchmark data and interpret it under a stated statistical discipline, in
which per-task decompositions locate where each model does its work,
small-stratum results are read by failure counts rather than rates, and
cross-model gaps that fall inside confidence intervals are treated as
ties rather than rankings. The verbal evidence has accumulated through
the placement, with benchmark results and design decisions presented at
the industry partner's weekly stand-ups and in regular meetings with the
academic supervisor, and concludes with the ESIPS oral presentation to
the academic supervisor and the industry partner, scheduled for
14 August 2026.